---
title: "Question 2"
author: "Trixie Grace Mok"
date: "2024-02-21"
output: html_document
---

```{r}
#load library
library(conflicted)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(scales)
library(mlr3)
library(mlr3learners)
library(glmnet)
library(mlr3pipelines)
library(mlr3tuning)
library(broom)
library(zoo)
library(pROC)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggpubr)
library(caret)
```
```{r}
# Set working directory
setwd("C:/Users/Trixie/Desktop/From WDBLACK/Year 2 UOL Subjects/Programming Assignment/Final/Question 2/dataverse_files")

```
```{r}
# Load data from csv files
airports <- read.csv("airports.csv")
carriers <- read.csv("carriers.csv")
planes <- read.csv("plane-data.csv")

flights_1999 <- read.csv("1999.csv.bz2")
flights_2000 <- read.csv("2000.csv.bz2")
flights_2001 <- read.csv("2001.csv.bz2")
flights_2002 <- read.csv("2002.csv.bz2")
flights_2003 <- read.csv("2003.csv.bz2")
flights_2004 <- read.csv("2004.csv.bz2")
flights_2005 <- read.csv("2005.csv.bz2")
flights_2006 <- read.csv("2006.csv.bz2")
flights_2007 <- read.csv("2007.csv.bz2")
flights_2008 <- read.csv("2008.csv.bz2")
```

```{r}
# Row bind the data from year 1999 to 2008
tenyears_flight <- rbind(flights_1999, flights_2000, flights_2001, flights_2002, flights_2003, flights_2004, flights_2005, flights_2006, flights_2007, flights_2008)
  
```


Question 2(a)
**What are the best times and days of the week to minimise delays each year?**

```{r}
best_time_plots <- list()

for (year in 1999:2008) {
  # retrieves the data frame for the current year
  year_flight <- get(paste0("flights_", year))
  
  # Clean data by filtering out cancelled and diverted flights, then remove duplicates
  year_flight_cleaned <- year_flight %>%
    filter(Cancelled == 0, Diverted == 0) %>%
    distinct()

  # Classify flights under 2-hour constant time frames
  year_flight_time_intervals <- year_flight_cleaned %>%
    mutate(dt_intervals = case_when(
      CRSDepTime >= 0 & CRSDepTime < 200 ~ "00:00-01:59",
      CRSDepTime >= 200 & CRSDepTime < 400 ~ "02:00-03:59",
      CRSDepTime >= 400 & CRSDepTime < 600 ~ "04:00-05:59",
      CRSDepTime >= 600 & CRSDepTime < 800 ~ "06:00-07:59",
      CRSDepTime >= 800 & CRSDepTime < 1000 ~ "08:00-09:59",
      CRSDepTime >= 1000 & CRSDepTime < 1200 ~ "10:00-11:59",
      CRSDepTime >= 1200 & CRSDepTime < 1400 ~ "12:00-13:59",
      CRSDepTime >= 1400 & CRSDepTime < 1600 ~ "14:00-15:59",
      CRSDepTime >= 1600 & CRSDepTime < 1800 ~ "16:00-17:59",
      CRSDepTime >= 1800 & CRSDepTime < 2000 ~ "18:00-19:59",
      CRSDepTime >= 2000 & CRSDepTime < 2200 ~ "20:00-21:59",
      CRSDepTime >= 2200 & CRSDepTime <= 2400 ~ "22:00-23:59",
      TRUE ~ "Unknown" # Handles edge cases or invalid data
    ))

  # Fill missing values in delay columns with last observation carried forward
  delay_columns <- c("ArrDelay", "DepDelay", "CarrierDelay", "NASDelay", "SecurityDelay", "LateAircraftDelay", "WeatherDelay")
  year_flight_filled <- year_flight_time_intervals
  for (col in delay_columns) {
    year_flight_filled[[col]] <- na.locf(year_flight_filled[[col]], na.rm = FALSE)
  }

  # Calculate total delay and filter flights with any delay
  year_flight_delays <- year_flight_filled %>%
    mutate(total_delay = rowSums(select(., all_of(delay_columns)), na.rm = TRUE)) %>%
    filter(total_delay > 0)

  # Aggregate average delay by time intervals
  best_time_day <- year_flight_delays %>%
    group_by(dt_intervals) %>%
    summarise(avg_delay = mean(total_delay, na.rm = TRUE))

  # Plot average delay for each time interval
  best_time_plots[[paste0("Year_", year)]] <- ggplot(best_time_day, aes(x = dt_intervals, y = avg_delay)) +
    geom_bar(stat = "identity", color = "black", fill = "pink", width = 0.8) +
    geom_text(aes(label = sprintf("%0.2f", avg_delay)), vjust = -0.4, size = 3) +
    labs(title = paste0("Average Delay by Departure Time Interval in ", year),
         x = "Departure Time Interval",
         y = "Average Delay (Mins)") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
}
# Display plots
best_time_plots
```


```{r}

# Initialise list to store plots

list_best_day <- list()

for (year in 1999:2008) {
  year_flight <- get(paste0("flights_", year))
  year_flight_cleaned <- year_flight %>%
    filter(Cancelled == 0, Diverted == 0) %>%
    distinct()
  
  delay_columns <- c("ArrDelay", "DepDelay", "CarrierDelay", "NASDelay", "SecurityDelay", "LateAircraftDelay", "WeatherDelay")
  year_flight_filled <- year_flight_cleaned
  for (col in delay_columns) {
    year_flight_filled[[col]] <- na.locf(year_flight_filled[[col]], na.rm = FALSE)
  }
  

  year_flight_delays <- year_flight_filled %>%
    mutate(total_delay = rowSums(select(., all_of(delay_columns)), na.rm = TRUE)) %>%
    filter(total_delay > 0)

    
  # Group to find best day of the week
  best_day_week <- year_flight_delays %>%
    filter(total_delay > 0) %>%
    group_by(DayOfWeek) %>% 
    summarise(avg_delay = mean(total_delay)) %>%
    arrange(DayOfWeek)
  best_day_week
  
    
  # Plot graph for best day of the week
    list_best_day[[paste0("Year_", year)]] <-ggplot(best_day_week, aes(x = factor(DayOfWeek, levels = c("1", "2", "3", "4", "5", "6", "7")),
                               y = avg_delay, label = sprintf("%0.2f", round(avg_delay, digits = 2)))) +
    geom_bar(stat = "identity", color = "black", fill = "pink", width = 0.8) +
    scale_x_discrete(labels = c("1" = "Monday", "2" = "Tuesday", "3" = "Wednesday", "4" = "Thursday", "5" = "Friday", "6" = "Saturday", "7" = "Sunday")) +
    labs(title = paste("Best Day Of Week in", year), x = "Day", y = "Average Delay (Mins)") + 
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.x = element_text(angle = 20, hjust = 1)) +
    geom_text(size = 4, hjust = 0.5, vjust = 1.5)

}
list_best_day
```

```{r}

# Initialise list to store plots
list_month_day <- list()

for (year in 1999:2008) {

  year_flight <- get(paste0("flights_", year))
  
  year_flight_cleaned <- year_flight %>%
    filter(Cancelled == 0, Diverted == 0) %>%
    distinct()
  
  delay_columns <- c("ArrDelay", "DepDelay", "CarrierDelay", "NASDelay", "SecurityDelay", "LateAircraftDelay", "WeatherDelay")
  for (col in delay_columns) {
    year_flight_cleaned[[col]] <- na.locf(year_flight_cleaned[[col]], na.rm = FALSE)
  }
  
  year_flight_delays <- year_flight_cleaned %>%
    mutate(total_delay = rowSums(select(., all_of(delay_columns)), na.rm = TRUE)) %>%
    filter(total_delay > 0)

  # Group to find average delay by month
  best_month_year <- year_flight_delays %>%
    group_by(Month) %>% 
    summarise(avg_delay = mean(total_delay), .groups = 'drop') %>%
    right_join(all_months_df, by = "Month") %>%
    arrange(Month)
  
  # Use last observation carried forward to fill missing avg_delay values
  best_month_year$avg_delay <- na.locf(best_month_year$avg_delay, na.rm = FALSE)

  
  # Plot graph for each month of the year
  list_month_day[[paste0("Year_", year)]] <- ggplot(best_month_year, aes(x = factor(Month, levels = 1:12),
                               y = avg_delay, label = sprintf("%0.2f", round(avg_delay, digits = 2)))) +
    geom_bar(stat = "identity", color = "black", fill = "pink", width = 0.8) +
    scale_x_discrete(labels = c("1" = "January", "2" = "February", "3" = "March", "4" = "April", "5" = "May", "6" = "June", 
                                "7" = "July", "8" = "August", "9" = "September", "10" = "October", "11" = "November", "12" = "December")) +
    labs(title = paste("Best Time Of Year", year), x = "Month", y = "Average Delay (Mins)") + 
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_text(size = 3, hjust = 0.5, vjust = 1.5) +
    theme(axis.text.x = element_text(angle = 30))
}

list_month_day

```


Question 2(b)
**Evaluate whether older planes suffer more delays on a year-to-year basis**


```{r}
for (year in 1999:2008) {

  year_flight <- get(paste0("flights_", year))
  
  year_flight_cleaned <- year_flight %>%
    filter(Cancelled == 0, Diverted == 0) %>%
    distinct()
  
  year_flight_cleaned <- year_flight_cleaned %>%
    mutate(date = make_date(Year, Month, DayofMonth))
  
  delay_columns <- c("ArrDelay", "DepDelay", "CarrierDelay", "NASDelay", "SecurityDelay", "LateAircraftDelay", "WeatherDelay")
  year_flight_cleaned[delay_columns] <- lapply(year_flight_cleaned[delay_columns], function(x) as.numeric(as.character(x)))
  
  # Fill missing values with the last observation carried forward
  year_flight_cleaned <- year_flight_cleaned %>%
    fill(everything(), .direction = "down")
  
  # calculate total_delay
  year_flight_cleaned <- year_flight_cleaned %>%
    mutate(total_delay = rowSums(select(., delay_columns), na.rm = TRUE)) %>%
    filter(total_delay > 0)

  planes <- mutate(planes, issue_date = as.Date(issue_date, format = "%m/%d/%Y"))

  planes_manu <- inner_join(year_flight_cleaned, planes, by = c("TailNum" = "tailnum"))
  
  planes_manu <- planes_manu %>%
    mutate(AircraftAge = as.integer((date - issue_date) / dyears(1)))
  
  planes_manu <- planes_manu %>%
    group_by(TailNum) %>%
    mutate(AvgDelay = mean(total_delay, na.rm = TRUE)) %>%
    ungroup()
  
  planes_manu_filtered <- filter(planes_manu, AircraftAge >= 0 & AircraftAge <= 22)
  
  p <- ggplot(planes_manu_filtered, aes(x = AircraftAge, y = AvgDelay)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    labs(title = paste("Average Delay vs. Aircraft Age for", year),
         x = "Aircraft Age (years)", y = "Average Delay (minutes)") +
    theme_minimal()
  
  print(p)
}

```


Question 2(c)
**For each year, fit a logistic regression model for the probability of diverted US flights using as many features as possible from attributes of the departure date, the scheduled departure and arrival times, the coordinates and distance between departure and planned arrival airports, and the carrier. Visualize the coefficients across years.**


```{r}
# Initialize a list to store coefficients for each year
coefficients_list <- list()

for (year in 1999:2008) {
  # Load the dataset for the current year
  year_flight <- get(paste0("flights_", year))
  
  # Sample 10,000 flights from the current year's data
  sample_fd <- year_flight %>%
    sample_n(size = 10000)
  
  # Join with 'airports' data to get coordinates for origin and destination airports
  # Select for logistic regression 
  year_flight_processed <- sample_fd %>%
    left_join(airports, by = c("Origin" = "iata")) %>%
    rename(Origin_lat = lat, Origin_long = long) %>%
    left_join(airports, by = c("Dest" = "iata")) %>%
    rename(Dest_lat = lat, Dest_long = long) %>%
    select(Year, Month, DayofMonth, CRSDepTime, CRSArrTime, UniqueCarrier, Distance, Diverted,
           Origin_lat, Origin_long, Dest_lat, Dest_long) %>%
    mutate(
      UniqueCarrier = as.factor(UniqueCarrier),
      Year = as.factor(Year),
      Month = as.factor(Month),
      Diverted = as.factor(Diverted),
      DayofMonth = as.factor(DayofMonth),
      CRSDepTime = as.factor(CRSDepTime),
      CRSArrTime = as.factor(CRSArrTime),
      Distance = as.numeric(as.character(Distance)),
      Origin_lat = as.numeric(as.character(Origin_lat)),
      Origin_long = as.numeric(as.character(Origin_long)),
      Dest_lat = as.numeric(as.character(Dest_lat)),
      Dest_long = as.numeric(as.character(Dest_long))
    ) %>%
    na.omit()

 factor_vars <- sapply(year_flight_processed, is.factor)
    single_level_factors <- sapply(year_flight_processed[, factor_vars], function(x) nlevels(x) < 2)
    
    # formula to derive factors with two or more levels, then another formula to use Diverted as the response variable for prediction which then takes the previous formula to select only factors with levels greater than 1
    formula_vars <- names(year_flight_processed)[!single_level_factors]
    formula_str <- paste("Diverted ~", paste(formula_vars[!formula_vars %in% "Diverted"], collapse = " + "))
    formula <- as.formula(formula_str)
    
    # Fit the logistic regression model using the updated formula
    logistic_yearly <- glm(formula, data = year_flight_processed, family = binomial())
    

  # Store model coefficients
  coefficients_list[[as.character(year)]] <- summary(logistic_yearly)$coefficients
}

# Prepare data for visualization
coef_name <- "Distance"
coef_values <- sapply(coefficients_list, function(coef) {
  if(coef_name %in% rownames(coef)) coef[coef_name, "Estimate"] else NA
})

years <- names(coefficients_list)
data_to_plot <- data.frame(Year = as.numeric(years), Coefficient = coef_values, stringsAsFactors = FALSE)

# Filter out years where the coefficients are missing 
data_to_plot <- na.omit(data_to_plot)

# Visualization
ggplot(data_to_plot, aes(x = Year, y = Coefficient)) +
  geom_line() +
  geom_point() +
  labs(title = paste("Evolution of", coef_name, "Coefficient over Years"),
       x = "Year", y = "Coefficient Estimate") +
  theme_minimal()


```


```{r}
for (year in 1999:2008) {
  # Load the dataset for the current year
  year_flight <- get(paste0("flights_", year))
  
  # Sample 10,000 flights from the current year's data
  sample_fd <- year_flight %>%
    sample_n(size = 10000)
  
  # Join with 'airports' data to get coordinates for origin and destination airports
  # Then select and mutate as necessary
  year_flight_processed <- sample_fd %>%
    left_join(airports, by = c("Origin" = "iata")) %>%
    rename(Origin_lat = lat, Origin_long = long) %>%
    left_join(airports, by = c("Dest" = "iata")) %>%
    rename(Dest_lat = lat, Dest_long = long) %>%
    select(Year, Month, DayofMonth, CRSDepTime, CRSArrTime, UniqueCarrier, Distance, Diverted,
           Origin_lat, Origin_long, Dest_lat, Dest_long) %>%
    mutate(
      UniqueCarrier = as.factor(UniqueCarrier),
      Year = as.factor(Year),
      Month = as.factor(Month),
      Diverted = as.factor(Diverted),
      DayofMonth = as.factor(DayofMonth),
      CRSDepTime = as.factor(CRSDepTime),
      CRSArrTime = as.factor(CRSArrTime),
      Distance = as.numeric(as.character(Distance)),
      Origin_lat = as.numeric(as.character(Origin_lat)),
      Origin_long = as.numeric(as.character(Origin_long)),
      Dest_lat = as.numeric(as.character(Dest_lat)),
      Dest_long = as.numeric(as.character(Dest_long))
    ) %>%
    na.omit()

 factor_vars <- sapply(year_flight_processed, is.factor)
    single_level_factors <- sapply(year_flight_processed[, factor_vars], function(x) nlevels(x) < 2)
    
    # formula excluding these factors
    formula_vars <- names(year_flight_processed)[!single_level_factors]
    formula_str <- paste("Diverted ~", paste(formula_vars[!formula_vars %in% "Diverted"], collapse = " + "))
    formula <- as.formula(formula_str)
    
    # Fit the logistic regression model using the updated formula
    logistic_yearly <- glm(formula, data = year_flight_processed, family = binomial())
    
 # Making predictions for the current year's data
  predictions <- predict(logistic_yearly, type = "response", newdata = year_flight_processed)
  
  # Calculate and plot ROC curve for the current year
  roc_result <- roc(response = year_flight_processed$Diverted, predictor = predictions)
  plot(roc_result, main = sprintf("ROC Curve for Year %d", year))
  auc_value <- auc(roc_result)
  print(sprintf("AUC for Year %d: %f", year, auc_value))
}
```


```{r}

# Specified predictors, excluding 'Diverted' as it's the outcome variable
predictors <- c("Year", "Month", "DayofMonth", "CRSDepTime", "CRSArrTime", "UniqueCarrier", "Distance", "Origin_lat", "Origin_long", "Dest_lat", "Dest_long")

# Initialize a list to store aggregated coefficient values for each predictor
coef_values_list <- list()

# Loop through each predictor to aggregate its coefficients across all years
for (coef_name in predictors) {
  coef_values_list[[coef_name]] <- sapply(coefficients_list, function(coef) {
    if (coef_name %in% rownames(coef)) coef[coef_name, "Estimate"] else NA
  })
}

# Convert the list of aggregated coefficients into a data frame
years <- 1999:2008  
coef_data <- do.call(rbind, lapply(names(coef_values_list), function(var) {
  data.frame(Variable = var, Year = years, Estimate = coef_values_list[[var]], stringsAsFactors = FALSE)
}))

# Year is numeric for plotting
coef_data$Year <- as.numeric(as.character(coef_data$Year))

# visualisation
ggplot(coef_data, aes(x = Year, y = Estimate, color = Variable)) +
  geom_line() +  
  geom_point() +  
  labs(title = "Evolution of Regression Coefficients Over Time",
       x = "Year", y = "Coefficient Estimate") +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "right")  


```
```{r}

predictors <- c("Year", "Month", "DayofMonth", "CRSDepTime", "CRSArrTime", "UniqueCarrier", "Distance", "Origin_lat", "Origin_long", "Dest_lat", "Dest_long")

# Loop through each predictor and create a plot
for (coef_name in predictors) {
  # Filter the data for the current predictor
  coef_values <- sapply(coefficients_list, function(coef) {
  if(coef_name %in% rownames(coef)) coef[coef_name, "Estimate"] else NA
})
  # Plot
  
  years <- names(coefficients_list)
  data_to_plot <- data.frame(Year = as.numeric(years), Coefficient = coef_values, stringsAsFactors = FALSE)
  

  data_to_plot <- na.omit(data_to_plot)
  
  
# Visualization
 g<- ggplot(data_to_plot, aes(x = Year, y = Coefficient)) +
   geom_line() +
   geom_point() +
   labs(title = paste("Evolution of", coef_name, "Coefficient over Years"),
       x = "Year", y = "Coefficient Estimate") +
   theme_minimal()
  print(g)
}

```


```{r}
summary(logistic_yearly)
# The Coefficients are not ideal predictors because their P values are all high. Thus it may not be ideal to plot them. However I have provided visualisations in previous graphs. The coefficients in the following logistic_yearly output were only Month, DayofMonth, and CRSDepTime.
```
```{r}

summary_logistic <- summary(logistic_yearly)

summary_text <- capture.output(summary_logistic)


print(tail(summary_text, 14))
```




